---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please modify README.Rmd and run `pkgdown::build_site()` to update README.md -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

# synapser

The `synapser` package provides an interface to [Synapse](<http://www.synapse.org>), a collaborative workspace for reproducible data intensive research projects, providing support for:

* integrated presentation of data, code and text
* fine grained access control
* provenance tracking

The `synapser` package lets you communicate with the Synapse platform to create collaborative data analysis projects and access data using the R programming language. Other Synapse clients exist for [Python](http://docs.synapse.org/python), [Java](https://github.com/Sage-Bionetworks/Synapse-Repository-Services/tree/develop/client/synapseJavaClient>), and [the web browser](https://www.synapse.org).

## Installation

`synapser` is available as a ready-built package for Microsoft Windows and Mac OSX. For Linux systems, it is available to install from source. Please also check out our [System Dependencies article](articles/systemDependencies.html) for instructions on how to install system dependencies on Linux environments.

`synapser` can be installed or upgraded using the standard `install.packages()` command, adding the [Sage Bionetworks R Archive Network (RAN)](http://ran.synapse.org) to the repository list, e.g.:

```{r eval=F}
install.packages("synapser", repos=c("http://ran.synapse.org", "http://cran.fhcrc.org"))
```
Alternatively, edit your `~/.Rprofile` and configure your default repositories:
```{r eval=F}
options(repos=c("http://ran.synapse.org", "http://cran.fhcrc.org"))
```
after which you may run `install.packages` without specifying the repositories:
```{r eval=F}
install.packages("synapser")
```

If you have been asked to validate a release candidate, please use:

```{r eval=F}
install.packages("synapser", repos=c("http://staging-ran.synapse.org", "http://cran.fhcrc.org"))
```

## Usage

To get started, try logging into Synapse. If you don't already have a Synapse account, register [here](https://www.synapse.org/register):

```{r eval=F}
library(synapser)
synLogin()
```

Please visit the `synapser` [docs site](http://sage-bionetworks.github.io/synapser/articles/synapser.html) or view our vignettes for using the `synapser` package:

```{r eval=F}
browseVignettes(package="synapser")
```

### Usage Examples:

#### [knit2synapse](https://github.com/Sage-Bionetworks/knit2synapse) 
Knit RMarkdown files to Synapse wikis

#### [syndccutils](https://github.com/Sage-Bionetworks/syndccutils)
Code for managing data coordinating operations (e.g., development of the CSBC/PS-ON Knowledge Portal and individual Center pages) for Sage-supported communities through Synapse.
