---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please modify README.Rmd and run `rmarkdown::render("README.Rmd")` to update README.md -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

# synapser

The `synapser` package provides an interface to
[Synapse](<http://www.synapse.org>), a collaborative
workspace for reproducible data intensive research projects,
providing support for:

* integrated presentation of data, code and text
* fine grained access control
* provenance tracking

The `synapser` package lets you communicate with the cloud-hosted
Synapse service to access data and create shared data analysis projects from
within Python scripts or at the interactive Python console. Other Synapse clients
exist for [Python](http://docs.synapse.org/python),
[Java](https://github.com/Sage-Bionetworks/Synapse-Repository-Services/tree/develop/client/synapseJavaClient>),
and [the web browser](https://www.synapse.org). 

## Installation

`synapser` is available as a ready-built package. It can
be installed or upgraded using the standard `install.packages()` command, adding the Sage Bionetworks repository to the repo list, e.g.:

```{r eval=F}
install.packages("synapser", repos=c("https://sage-bionetworks.github.io/ran", "http://cran.fhcrc.org"))
```
Alternatively, configure your default repo's in your `~/.Rprofile` like so:
```{r eval=F}
options(repos=c("https://sage-bionetworks.github.io/ran", "http://cran.fhcrc.org"))
```
after which you may run install.packages without specifying the repositories:
```{r eval=F}
install.packages("synapser")
library(synapser)
synLogin()
```

If you have been asked to validate a release candidate, please replace the URL https://sage-bionetworks.github.io/ran with https://sage-bionetworks.github.io/staging-ran, that is:

```{r eval=F}
install.packages("synapser", repos=c("https://sage-bionetworks.github.io/staging-ran", "http://cran.fhcrc.org"))
```

## Usage

Please visit our [docs site](http://docs.synapse.org/r/) or view our vignettes for using synapser package:

```{r eval=F}
browseVignettes(package="synapser")
```