---
output: github_document
---


```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

# synapser

The `synapser` package provides an interface to
[Synapse](http://www.synapse.org), a collaborative workspace for
reproducible data intensive research projects, providing support for:

* integrated presentation of data, code and text
* fine grained access control
* provenance tracking

The `synapser` package lets you communicate with the Synapse platform to
create collaborative data analysis projects and access data using the R
programming language. Other Synapse clients exist for
[Python](http://docs.synapse.org/python),
[Java](https://github.com/Sage-Bionetworks/Synapse-Repository-Services/tree/develop),
and [the web browser](https://www.synapse.org).

## Installation

`synapser` is available as a ready-built package for Microsoft Windows
and Mac OSX. For Linux systems, it is available to install from source.
Please also check out our [System Dependencies
article](https://r-docs.synapse.org/articles/systemDependencies.html) for instructions on how to
install system dependencies on Linux environments.

## How to Upgrade Python on Windows

- Download the Python installer from the Official Website of Python
[here](https://www.python.org/downloads/windows/).
- Install the Downloaded Python Installer
- check Install Python and Check the “Add python.ext to PATH”, then click on the “Install Now” button.
- Verify the Update

  ```{cmd eval=F}
  python --version
  ```

- `Note` If it still shows the old version, you may restart your system. Or uninstall the old version from the control panel.

## How to Upgrade Python on macOS

- Both python 2x and 3x can stay installed in a MAC. Mac comes with python 2x version. To check the default python version in your MAC, open the terminal and type

  ```{cmd eval=F}
  python --version
  Python3 --version
  ```

- If you don't then go ahead and install it with the installer. Go the the python's official site
[here](https://www.python.org/downloads/mac-osx/).

- Now restart the terminal and check again with both commands python —version

  ```{cmd eval=F}
  Python3 --version
  ```

- `Or` use to `install last version`

  ```{cmd eval=F}
  brew install python3 && cp /usr/local/bin/python3 /usr/local/bin/python
  ```

## How to Upgrade Python on Linux

- Add the repository and update

  ```{cmd eval=F}
  sudo add-apt-repository ppa:deadsnakes/ppa
  sudo apt-get update
  ```

- Update the package list

  ```{cmd eval=F}
  apt-get update
  ```

- Verify the updated Python packages list

    ```{cmd eval=F}
    $apt list | grep python3.10
    ```

- Install the Python 3.10 package using apt-get

    ```{cmd eval=F}
    sudo apt-get install python3.10
    ```

- Add Python 3.8 & Python 3.10 to update-alternatives

  ```{cmd eval=F}
  sudo update-alternatives --install /usr/bin/python3 python3 /usr/bin/python3.8 1
  sudo update-alternatives --install /usr/bin/python3 python3 /usr/bin/python3.10 2
  ```

- Update Python 3 for point to Python 3.10

  ```{cmd eval=F}
  sudo update-alternatives --config python3
  ```

## How to Upgrade R on Windows

- Verify R version
  - Open RStudio > At the top of the Console you will see session info > The first line tells you which version of R you are using.
  - `or` write in console >'R.version.string' to print out the R version.
  - go to Tools > Check for Package Updates. If there’s an update available for tidyverse, install it.

## for Windows

- To update R on Windows, try using the package installer (only for Windows).
- Got to Tools (at the top) > Check for package updates. If tidyverse shows up on the list, select it, then click “Install Updates.”

## for Mac

- Go to
[here](https://cloud.r-project.org/bin/macosx/).
- Click the link you need to update R pkg
- When the file finishes downloading, double-click to install. You should be able to click “Next” to all dialogs to finish the installation.
- From within RStudio, go to Help > Check for Updates to install newer version of RStudio (if available, optional).
- To update packages, go to Tools > Check for Package Updates. If updates are available, select All (or just tidyverse), and click Install Updates.

## To build and install follow these steps

- build package from R console:

  ```{r eval=F}
  R CMD BUILD .
  ```

- install package from R console:

  ```{r eval=F}
  R CMD INSTALL synapser_0.1.0.tar.gz
  ```

## Usage

To get started, try logging into Synapse. If you don’t already have a
Synapse account, register [here](https://www.synapse.org/register):

```{r eval=F}
library(synapser)
synLogin()
```

Please visit the `synapser` [docs site](http://sage-bionetworks.github.io/synapser/articles/synapser.html) or view our vignettes for using the `synapser` package:

```{r eval=F}
browseVignettes(package="synapser")
```

### Usage Examples

#### [knit2synapse](https://github.com/Sage-Bionetworks/knit2synapse)

Knit RMarkdown files to Synapse wikis

#### [syndccutils](https://github.com/Sage-Bionetworks/syndccutils)

Code for managing data coordinating operations (e.g., development of the CSBC/PS-ON Knowledge Portal and individual Center pages) for Sage-supported communities through Synapse.
