---
title: "Authentication"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Manage Synapse Credentials}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

There are multiple ways one can login to Synapse. We recommend users choose the method that fits their workflow best.

## Prerequisites

* Create a [Personal Access Token](https://help.synapse.org/docs/Managing-Your-Account.2055405596.html#ManagingYourAccount-PersonalAccessTokens) (**aka: Synapse Auth Token**) obtained
from synapse.org under your Settings.
    * Note that a token must minimally have the **view** scope to be used with the Synapse R Client.
    * Include **Download** and **Modify** permissions if you are using the Synapse R Client to follow any subsequent tutorials.
* Once a personal access token has been created it can be used for any of the options below.

## One Time Login

Use the `synLogin()` function.

```{r eval=FALSE}
library(synapser)
synLogin(authToken = "authtoken")
# Prints: Welcome, First Last!
```

## Use `.synapseConfig`

For writing code using the Synapse R Client (`synapser`) that is easy to share with others, please do not include your credentials in the code. Instead, please use the `~/.synapseConfig` file to manage your credentials.

The Synapse R Client supports multiple profiles within the `~/.synapseConfig` file, enabling users to manage credentials for multiple accounts. Each profile is defined in its own `[profile <profile_name>]` section. A default profile can still be defined using `[default]`.

### Editing the `~/.synapseConfig` file

Open the `~/.synapseConfig` file using your preferred text editing tool and find/insert the following section(s):

```
[default]
username = default_user
authtoken = default_auth_token

[profile user1]
username = user1
authtoken = user1_auth_token

[profile user2]
username = user2
authtoken = user2_auth_token

# This section is deprecated. It will be used if a `default` profile or a specific profile is not present in the config file
#[authentication]
#username = default_user
#authtoken = default_auth_token
```

`username` is optional when using `authtoken`, but if provided, an additional check to verify the `authtoken` matches the `username` is performed.

The `authtoken` is also known as a personal access token. It is generated from your [synapse.org Settings](https://help.synapse.org/docs/Managing-Your-Account.2055405596.html#ManagingYourAccount-PersonalAccessTokens).

### Transitioning from One Profile to Multiple

If you're currently using a single profile (under the `[default]` or `[authentication]` section) and wish to start using multiple profiles,
simply add new sections for each profile with a unique profile name. For example, you can add a profile for user1 and user2 as shown below.
The Synapse R Client will allow you to choose which profile to use at login.

```
[default]
username = default_user
authtoken = default_auth_token

[profile user1]
username = user1
authtoken = user1_auth_token

[profile user2]
username = user2
authtoken = user2_auth_token
```

## Logging in with your ~/.synapseConfig file

**Note:** If no profile is specified the `default` section will be used. Additionally, to support backwards compatibility, `authentication` will continue to function. `authentication` will be used if no profile is used and `default` is not present in the configuration file.

### Logging in via R

```{r eval=FALSE}
library(synapser)
synLogin()
```

If you want to log in with a specific profile, simply pass the profile name as an argument to `synLogin()`:

```{r eval=FALSE}
synLogin(profile = "user1")
```


### For more information, see:

- [synLogin](../reference/synLogin.html)
- [synLogout](../reference/synLogout.html)
- [synapse config](configuration.html)
