---
title: "Data Upload, Download, and Delete"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Data Upload, Download, and Delete}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Overview

This instruction demonstrates how to upload, download, and delete data using `synapser`.


Key differences from the legacy API:

| Task | Legacy | New pipe UI |
|---|---|---|
| Store an entity | `synStore(obj, ...)` | `File(...) |> synStore(file_options, container_options, ...)` or `synStore(entity = File(...), parent, file_options, container_options, ...)`|
| Suppress new version | `synStore(obj, forceVersion = FALSE)` | `File(...) |> synStore()` |
| Set version label | `synStore(obj, versionLabel = "2")` | `synGet(id) |> (\(x) {x$version_label <- "2"; x})() |> synStore()` |
| Track provenance | `synStore(obj, used = "syn123", executed = "syn456")` | `synGet(id) |> (\(x) {x$activity <- Activity(used = list("syn123"), executed = list("syn456")); x})() |> synStore()` |
| Access entity ID | `entity$properties$id` | `entity$id` |
| Get an entity | `synGet(entity, ...)` | `File(...) |> synGet(file_options, ...)` or `synGet(synapse_id, file_options, ...)` |
| Get a specific version | `synGet(entity, version = 1)` | `File(id, version_label = 1) |> synGet()` or `File(id) |> synGet(version_number = 1)` or `synGet(synapse_id, version_number = 1)` |
| Skip file download | `synGet(entity, downloadFile = FALSE)` | `File(...) |> synGet(file_options = FileOptions(download_file = FALSE))` or `synGet(synapse_id, file_options = FileOptions(download_file = FALSE))` |
| Delete a version | `synDelete(obj, version = 1)` | `File(...) |> synDelete(version = 1,  version_only = TRUE)` or `synDelete(entity, version = 1,  version_only = TRUE)` |

## Load the package
```{r eval=FALSE}
library(synapser)
```

## Login

```{r eval=FALSE}
synLogin()
```

Credentials are read from `~/.synapseConfig` or the `SYNAPSE_AUTH_TOKEN`
environment variable. See the
[Managing Credentials](manageSynapseCredentials.html) vignette for setup
details.

## Create a Project

```{r eval=FALSE}
project <- Project(name="My uniquely named project about Alzheimer's Disease_danlu_r") |> synStore()
cat("Created project:", project$id, "\n")
```

`synStore()` returns the stored entity with the assigned Synapse ID in
`project$id`. In the new models API, IDs are accessed directly on the object
(`project$id`) rather than through `project$properties$id`.

## Create a Folder

Create a folder inside the project:

```{r eval=FALSE}
data_folder <- Folder(name = "single_cell_RNAseq_batch_1", parent_id = project$id) |> synStore()
cat("Created folder:", data_folder$id, "\n")
```

## Upload a File

Create a local file and upload it to the folder. 

```{r eval=FALSE}
file <- File(path = "~/my_ad_project/biospecimen_experiment_1/fileA.txt", parent_id = data_folder$id) |> synStore()
cat("Uploaded file:", file$id, "\n")
```
Storing the same file path as an existing version does not create a duplicate. Synapse detects unchanged content by MD5.

## Version the file

```{r eval=FALSE}
file <- synGet(synapse_id = file$id) |>
    (\(x) {
      x$version_label   <- "2"
      x$version_comment <- "Added version 2"
      x
    })() |>
    synStore()
```

## Add or Update Annotations

Use `synStore()` to add or update annotations.
```{r eval=FALSE}
file$annotations  <- list(assay = "RNA-seq", fileType = "csv", dataStage = "processed")
file <- file |> synStore()
```

## Download a File

Retrieve a file by its Synapse ID. The file is downloaded to the local cache `.synapseCache` by default. 
To set a download path, use `FileOptions(download_location = "...")` in the `file_options` parameter. 

```{r eval=FALSE}
downloaded_file <- synGet(file$id, file_options = FileOptions(download_location = "..."))
cat("Downloaded to:", downloaded_file$path, "\n")
```

## Download a Specific Version

Use `version_number` to pin to a particular version:

```{r eval=FALSE}
version_1 <- synGet(file$id, version_number = 1)
cat("Version 1 path:", version_1$path, "\n")
```

## Clean Up

Delete the project and all its contents when done:

```{r eval=FALSE}
synDelete(project)
```
