---
title: "File View"
author: "KimYen Ladia"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{File View}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## File View

A file view is a view of all Files within one or more Projects or Folders. File views can:

* Provide a way of isolating or linking data based on similarities
* Provide the ability to link Files together by their annotations
* Allow view/editing File attributes in bulk
* Allow Files to be easily searched and queried

Preliminaries:
```{r collapse=TRUE}
library(synapser)
synLogin()
# Create a new project
# use hex_digits to generate random string
hex_digits <- c(as.character(0:9), letters[1:6])
projectName <- sprintf("My unique project %s", paste0(sample(hex_digits, 32, replace = TRUE), collapse = ""))
project <- Project(projectName)
project <- synStore(project)

# Create some files
filePath <- tempfile()
connection <- file(filePath)
writeChar("this is the content of the first file", connection, eos = NULL)
close(connection)
file <- File(path = filePath, parent = project)
file <- synStore(file)

filePath2 <- tempfile()
connection2 <- file(filePath2)
writeChar("this is the content of the second file", connection, eos = NULL)
close(connection2)
file2 <- File(path = filePath2, parent = project)
file2 <- synStore(file2)

# add some annotations
synSetAnnotations(file, list(contributor = "Sage", class = "V"))
synSetAnnotations(file2, list(contributor = "UW", rank = "X"))
```

Creating a File View:
```{r collapse=TRUE}
view <- EntityViewSchema(name = "my first file view",
                         columns = c(
                           Column(name = "contributor", columnType = "STRING"),
                           Column(name = "class", columnType = "STRING"),
                           Column(name = "rank", columnType = "STRING")),
                         parent = project$properties$id,
                         scopes = project$properties$id,
                         add_default_columns = TRUE)

view <- synStore(view)
```

```{r include = FALSE}
# wait for the view to be created
Sys.sleep(10)
```

To see the content of your newly created File View, use synTableQuery():
```{r collapse=TRUE}
queryResults <- synTableQuery(sprintf("select * from %s", view$properties$id))
data <- as.data.frame(queryResults)
data
```

## Updating Annotations using File View

To update 'class' annotation for 'file2', simply update the view:
```{r collapse=TRUE, eval=F}
data["class"] <- c("V", "VI")
synStore(Table(view$properties$id, data))
```

The change in annotations reflect in synGetAnnotations():
```{r collapse=TRUE, eval=F}
synGetAnnotations(file2$properties$id)
```

A File View is a Table. Please visit [Tables vignettes](table.html) to see how to change schema, update content, and other operations that can be done on File View.

```{r collapse=TRUE}
synDelete(project)
```
