---
title: "File View"
author: "KimYen Ladia"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{File View}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## File View

A file view is a view of all Files within one or more Projects or Folders. File views can:

* Provide a way of isolating or linking data based on similarities
* Provide the ability to link Files together by their annotations
* Allow view/editing File attributes in bulk
* Allow Files to be easily searched and queried

Preliminaries:
```{r collapse=TRUE}
library(synapser)
synLogin()
# Create a new project
projectName<-sprintf("My unique project created on %s", format(Sys.time(), "%a %b %d %H%M%OS4 %Y"))
project<-Project(projectName)
project<-synStore(project)
# Create some file
filePath<- tempfile()
connection<-file(filePath)
writeChar("this is the content of the file", connection, eos=NULL)
close(connection)  
file <- File(path=filePath, parent=project)
file <- synStore(file)
```
Creating a File View:

```{r collapse=TRUE}
view <- EntityViewSchema(name='my first file view',
                         parent=project$properties$id,
                         scopes=c(project$properties$id))

view <- synStore(view)
```
To see the content of your newly created File View, use synTableQuery():
```{r collapse=TRUE}
query_results <- synTableQuery(sprintf("select * from %s", view$properties$id))
data <- query_results$asDataFrame()
data
```
A File View is a Table. Please visit [Tables vignettes](table.html) to see how to change schema, update content, and other operations that can be done on File View.

```{r collapse=TRUE}
synDelete(project)
```
