---
title: "Manage Synapse Credentials"
author: "KimYen Ladia"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Manage Synapse Credentials}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
## Manage Synapse Credentials

There are multiple ways one can login to Synapse. We recommend users to choose the method that fits your workflow.

### One Time Login

Use username and password to login as follow:
```{r collapse = TRUE, eval = F}
library("synapser")
synLogin("username", "password")
```

### Use `.synapseConfig`

For writing code using the `synapser` client, that easy to share with others, please do not include your credentials in the code. Instead, please use `.synapseConfig` file. You can prefer this documentation to your collaborators on how to setup `.synapseConfig` file.

First, reate a `.synapseConfig` file in your home directory with the following content:
```{r collapse = TRUE, eval = F}
[authentication]
username = your_username
password = your_password

```

Then login without specifying your username and password:
```{r collapse = TRUE, eval = F}
library("synapser")
synLogin()
```

### Letting the Operating System Manage Your Synapse Credentials

For users who would like to save their credentials and let other OS configured application (like keychain in Mac) manages credentials for you, when login the first time, use:
```{r collapse = TRUE, eval = F}
library("synapser")
synLogin("username", "password", rememberMe=True)
```

Then the application (keychain in Mac) will prompt you to allow other applications including Python and R to access this credentials. Please choose "Yes" or "OK".

The second time you login, you will not have to enter your username and password:
```{r collapse = TRUE, eval = F}
library("synapser")
synLogin()
```
