---
title: "Submission Views"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Submission Views}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
## Submission Views

SubmissionViews allow you to aggregate and query submissions from one or more evaluation queues in a tabular format. These views are useful for managing submissions in challenges, tracking scoring progress, and analyzing submission data across different evaluation queues.

This vignette walks through the basics of working with SubmissionViews using synapser.

### Tutorial purpose

In this vignette you will:

1. Set up and create an evaluation queue
2. Create a SubmissionView for the evaluation queue
3. Create and submit a file to the evaluation queue
4. Query and update the submission status
5. Modify the SubmissionView scope
6. Create a snapshot of the view
7. Query the snapshot

### Prerequisites

1. This tutorial assumes that you have a Synapse project.
2. The name of the Evaluation must also be globally unique. Please update the name used below with a new value.

## 1. Setting up and creating an evaluation queue

Before creating a SubmissionView, log in to Synapse, retrieve your project, and create an evaluation queue that will be used in the view.

You will want to replace `"My uniquely named project about Alzheimer's Disease"` with the name of your project.

```{r collapse=TRUE}
library(synapser)
synLogin()

# Retrieve the project ID
project = synGet(entity_name = "My uniquely named project about Alzheimer's Disease", parent_id = NULL)
cat("My project ID is:", project$id, "\n")

suffix = paste(sample(c(letters, LETTERS, 0:9), 8, replace = TRUE), collapse = "")

evaluation = Evaluation(
  name = paste0("Test Evaluation Queue for Alzheimer conference ", suffix),
  description = "Evaluation queue for testing submission view",
  content_source = project$id,
  submission_instructions_message = "Please submit your file for review.",
  submission_receipt_message = "Thank you for your submission."
) |> synStore()
cat("Created evaluation queue with ID:", evaluation$id, "\n")
```

## 2. Creating a SubmissionView for the evaluation queue

Next, create a SubmissionView that includes the evaluation queue in its scope. Custom columns can be added for metrics that will be used to score submissions, and columns can be reordered for a more useful display.

```{r collapse=TRUE}
view = SubmissionView(
  name = "SubmissionView for Alzheimer conference",
  parent_id = project$id,
  scope_ids = list(evaluation$id),
  include_default_columns = TRUE,
  columns = list(
    Column(name = "metric_A", column_type = "DOUBLE"),
    Column(name = "metric_B", column_type = "DOUBLE")
  ),
  activity = Activity(
    name = "Submission Review Analysis",
    description = "Analysis of Q1 2025 challenge submissions",
    used = list(
      UsedURL(
        name = "Challenge Homepage",
        url = "https://sagebionetworks.org/community/challenges-portal"
      )
    )
  )
) |> synStore()
cat("My SubmissionView ID is:", view$id, "\n")

# Reorder columns for better display
view |> synReorderColumn(name = "name", index = 2L)
view |> synReorderColumn(name = "status", index = 3L)
view |> synReorderColumn(name = "evaluationid", index = 4L)
view |> synStore()

cat("Available columns in the view:", names(view$columns), "\n")
```

## 3. Creating and submitting a file to the evaluation queue

Now create a test file and submit it to the evaluation queue. `tempfile()` is used for convenience — the file is written to a temporary location.

```{r collapse=TRUE}
temp_file_path = tempfile(fileext = ".txt")
writeLines("This is a test submission file.", temp_file_path)

# Upload the temporary file to Synapse
submission_file = File(
  path = temp_file_path, parent_id = project$id, name = "Test Submission"
) |> synStore()

# Submit the file to the evaluation queue
submission = Submission(
  entity_id = submission_file$id,
  evaluation_id = evaluation$id,
  name = "Test Submission"
) |> synStore()

cat("Created submission with ID:", submission$id, "\n")
```

## 4. Querying and updating the submission status

After submitting a file, you can query the SubmissionView to see the submission and update its status with scoring metrics.

Note: Due to Synapse's eventual consistency model, you may need to query more than once before the submission appears in the view.

```{r collapse=TRUE}
query = sprintf("SELECT * FROM %s WHERE id = '%s'", view$id, submission$id)
results_as_dataframe = synQuery(query)
# Due to the eventual consistency of the system, we perform a second query
results_as_dataframe = synQuery(query)

cat("Query results:\n")
results_as_dataframe

# Update the status to indicate it has been scored
submission_status = SubmissionStatus(id = submission$id) |> synGet()
cat("Submission status:", submission_status$status, "\n")

submission_status$status = "SCORED"
submission_status$submission_annotations[["metric_A"]] = list(90)
submission_status$submission_annotations[["metric_B"]] = list(80)
submission_status = submission_status |> synStore()
cat("Updated submission status to:", submission_status$status, "\n")
```

## 5. Modifying the SubmissionView scope

As your challenge evolves, you might need to add more evaluation queues to your SubmissionView. Here's how to create another evaluation queue and add it to your view's scope.

```{r collapse=TRUE}
# First let's make sure we have the latest view from Synapse:
view = SubmissionView(id = view$id) |> synGet()

# Create another evaluation queue to demonstrate adding to the scope
second_evaluation = Evaluation(
  name = paste0("Second Test Evaluation Queue for Alzheimer conference ", suffix),
  description = "Another evaluation queue for testing submission view",
  content_source = project$id,
  submission_instructions_message = "Please submit your file for review.",
  submission_receipt_message = "Thank you for your submission."
) |> synStore()
cat("Created second evaluation queue with ID:", second_evaluation$id, "\n")

# Add the new evaluation queue to the view's scope
view$scope_ids = append(view$scope_ids, second_evaluation$id)
view = view |> synStore()
cat("Updated SubmissionView scope. Current scope IDs:", unlist(view$scope_ids), "\n")
```

## 6. Creating a snapshot of the view

SubmissionViews support creating snapshots, which capture the state of all submissions at a specific point in time. This is useful for archiving or comparing submission states.

```{r collapse=TRUE}
snapshot = view |> synSnapshot(comment = "Initial submission review snapshot")
cat("Created snapshot of the SubmissionView:\n")
snapshot
```

## 7. Querying the snapshot

After creating a snapshot, you can query it to retrieve the state of submissions at the time the snapshot was created. This is useful for historical analysis or auditing.

The snapshot version is the version of the view immediately before the current one, i.e. the view's current `version_number` minus 1.

```{r collapse=TRUE}
view = SubmissionView(id = view$id) |> synGet()
snapshot_version = view$version_number - 1L

snapshot_query = sprintf("SELECT * FROM %s.%s", view$id, snapshot_version)
snapshot_results = synQuery(snapshot_query)
cat("Query results from the snapshot:\n")
snapshot_results
```
