---
title: "Submissions, SubmissionStatuses, SubmissionBundles"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Submissions, SubmissionStatuses, SubmissionBundles}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## What are Submissions?

In Synapse, a Submission is your entry to a challenge or evaluation queue. When you participate in a computational challenge or collaborative project, you submit your work (such as predictions, models, or analysis results) as a Submission to be evaluated and scored.

## Key Concepts

Before working with Submissions, it's helpful to understand how they fit into Synapse:

- **Entity**: Your actual work stored in Synapse (like a File containing predictions or a Docker image with your model)
- **Evaluation**: A queue that accepts and organizes submissions for a specific challenge or project
- **Submission**: The object associated with submitting your Entity to an Evaluation queue, creating a record that can be tracked and scored

## How Submissions Work

When you submit an Entity to an Evaluation:

- The Submission creates an immutable record linking your Entity to that Evaluation
- The Evaluation owns this Submission record (not you as the submitter)
- Organizers can add scores and feedback through a SubmissionStatus object
- You can track your Submissions and view their statuses

## Related Objects

synapser provides three object types for working with Submissions:

- **Submission**: Represents your entry in an Evaluation queue
- **SubmissionStatus**: Tracks scoring information and feedback for a Submission
- **SubmissionBundle**: Combines a Submission and its SubmissionStatus for convenient access

## What You'll Learn

This tutorial covers two perspectives:

1. **Participating in challenges**: Tracking and cancelling your submissions
1. **Organizing challenges**: Scoring and managing submissions from participants

## Tutorial Purpose

In this tutorial:

As a participant of a Synapse challenge, you will

1. Count the submissions in an existing evaluation queue
1. Fetch all of your submissions from an existing evaluation queue on Synapse
1. Cancel your submission

As an organizer of a Synapse challenge, you will

1. Batch-update submission statuses
1. Fetch the submission bundles for an evaluation queue

Making a submission, fetching a single Submission or SubmissionStatus by its ID, storing changes to a single SubmissionStatus, and deleting a Submission are not currently supported in synapser.

## Prerequisites

* You have an existing [Evaluation](evaluation.html) on Synapse to work from
* For the participant section, you have already made a submission to that Evaluation
* You have the correct permissions on the Evaluation queue for your desired tutorial section (participant or organizer)

## 1. Participating in a Synapse challenge

### Initial setup

```{r eval=FALSE}
library(synapser)
synLogin()

# REQUIRED: Set these to your actual Synapse IDs
# Do NOT leave these as NULL - the script will not work properly
EVALUATION_ID = NULL  # Replace with the evaluation queue ID you submitted to
ENTITY_ID = NULL  # Replace with the entity ID you want to submit

stopifnot(
  "EVALUATION_ID must be set to the evaluation queue ID" = !is.null(EVALUATION_ID),
  "ENTITY_ID must be set to the ID of a submission you made" = !is.null(ENTITY_ID)
)

print(sprintf("Working with Evaluation: %s", EVALUATION_ID))
print(sprintf("Submitting Entity: %s", ENTITY_ID))
```

### 1. Count your submissions

```{r eval=FALSE}
# Get the total count of submissions for this evaluation
submission_count = synGetSubmissionCount(evaluation_id = EVALUATION_ID)

print(submission_count)
```

### 2. Fetch all of your submissions from an existing evaluation queue on Synapse

```{r eval=FALSE}
# Get all of your submissions for this evaluation
user_submissions = as.list(Submission() |> synGetUserSubmissions(evaluation_id = EVALUATION_ID))

print(sprintf("Found %d submissions from the current user:", length(user_submissions)))
for (i in seq_along(user_submissions)) {
  sub = user_submissions[[i]]
  print(sprintf("  %d. ID: %s, Name: %s, Created: %s", i, sub$id, sub$name, sub$created_on))
}
```

### 3. Cancel your submission

Only cancel a submission if it allows it. The cancellation code is commented out by default; uncomment it if you want to test this functionality.

```{r eval=FALSE}
cancelled_submission = Submission(id = SUBMISSION_ID) |> synCancel()
print(sprintf("Submission %s has been requested for cancellation", cancelled_submission$id))
```

## 2. Organizing a Synapse challenge

### Initial setup

```{r eval=FALSE}
library(synapser)
synLogin()

# REQUIRED: Set these to your actual Synapse IDs
# Do NOT leave these as NULL - the script will not work properly
EVALUATION_ID = NULL  # Replace with the evaluation queue ID you submitted to
SUBMISSION_ID = NULL  # Replace with the ID of a submission you made

stopifnot(
  "EVALUATION_ID must be set to the evaluation queue ID" = !is.null(EVALUATION_ID),
  "SUBMISSION_ID must be set to the ID of a submission you made" = !is.null(SUBMISSION_ID)
)

print(sprintf("Working with Evaluation: %s", EVALUATION_ID))
print(sprintf("Working with Submission: %s", SUBMISSION_ID))
```

### 1. Batch-update submission statuses

```{r eval=FALSE}
# First, get all submission statuses that need updating
statuses_to_update = synGetAllSubmissionStatuses(
  evaluation_id = EVALUATION_ID,
  status = "RECEIVED"  # Get submissions that haven't been scored yet
)

print(sprintf("Found %d submissions to batch update", length(statuses_to_update)))

if (length(statuses_to_update) > 0) {
  # Update each status with validation information
  for (i in seq_along(statuses_to_update)) {
    status = statuses_to_update[[i]]
    status$status = "VALIDATED"
    status$submission_annotations = list(
      validation_status = list("PASSED"),
      validation_timestamp = list("2024-11-24T10:30:00Z"),
      batch_number = list(i),
      validator = list("automated_system")
    )
  }

  # Perform batch update
  batch_response = synBatchUpdateSubmissionStatuses(
    evaluation_id = EVALUATION_ID,
    statuses = statuses_to_update,
    is_first_batch = TRUE,
    is_last_batch = TRUE
  )

  print("Batch update completed successfully!")
  print(batch_response)
} else {
  print("No submissions found with 'RECEIVED' status to update")
}
```

### 2. Fetch the submission bundles for an evaluation queue

```{r eval=FALSE}
# Get all scored submission bundles for the evaluation
print("Fetching all submission bundles for the evaluation...")

bundles = as.list(SubmissionBundle() |> synGetEvaluationSubmissionBundles(
  evaluation_id = EVALUATION_ID,
  status = "SCORED"  # Only get scored submissions
))

print(sprintf("Found %d scored submission bundles", length(bundles)))

first_bundles = head(bundles, 5)  # Show first 5
for (i in seq_along(first_bundles)) {
  submission = first_bundles[[i]]$submission
  status = first_bundles[[i]]$submission_status

  print(sprintf("Bundle %d:", i))
  if (!is.null(submission)) {
    print(sprintf("  Submission ID: %s", submission$id))
    print(sprintf("  Submitter: %s", submission$submitter_alias))
    print(sprintf("  Entity ID: %s", submission$entity_id))
    print(sprintf("  Created: %s", submission$created_on))
  }

  if (!is.null(status)) {
    print(sprintf("  Status: %s", status$status))
    print(sprintf("  Modified: %s", status$modified_on))
    annotations = status$submission_annotations
    if (length(annotations) > 0) {
      print("  Scores:")
      for (key in names(annotations)) {
        if (key %in% c("accuracy", "f1_score", "precision", "recall")) {
          print(sprintf("    %s: %s", key, paste(unlist(annotations[[key]]), collapse = ", ")))
        }
      }
    }
  }
}
```

## References

- `Submission`
- `SubmissionStatus`
- `SubmissionBundle`
- `synLogin`
