---
title: "Synapse R Client Overview"
author: "Bruce Hoff"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Synapse R Client Overview}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


## Overview

The `synapser` package provides an interface to
[Synapse](<http://www.synapse.org>), a collaborative
workspace for reproducible data intensive research projects,
providing support for:

* integrated presentation of data, code and text
* fine grained access control
* provenance_ tracking

The `synapser` package lets you communicate with the cloud-hosted
Synapse service to access data and create shared data analysis projects from
within Python scripts or at the interactive Python console. Other Synapse clients
exist for [Python](http://docs.synapse.org/python),
[Java](https://github.com/Sage-Bionetworks/Synapse-Repository-Services/tree/develop/client/synapseJavaClient>),
and [the web browser](https://www.synapse.org). 

If you're just getting started with Synapse,
have a look at the [Getting Started guides for Synapse](http://docs.synapse.org/articles/getting_started.html)
.

Good example projects are:

* [TCGA Pan-cancer (syn300013)](https://www.synapse.org/#!Synapse:syn300013)
* [Development of a Prognostic Model for Breast Cancer Survival in an Open Challenge Environment (syn1721874)](https://www.synapse.org/#!Synapse:syn1721874)
* [Demo projects (syn1899339)](https://www.synapse.org/#!Synapse:syn1899339)

## Installation

`synapser` is available as a ready-built package. It can
be installed or upgraded using the standard `install.packages()` command, adding the Sage Bionetworks repository to the repo list, e.g.:

```{r eval=F}
install.packages("synapser", repos=c("https://sage-bionetworks.github.io/drat", "https://cran.cnr.berkeley.edu/"))
```
Alternatively, configure your default repo's in your `~/.Rprofile` like so:
```{r eval=F}
options(repos=c("https://sage-bionetworks.github.io/drat", "https://cran.cnr.berkeley.edu/"))
```
after which you may run install.packages without specifying the repositories:
```{r eval=F}
install.packages("synapser")
```

## Connecting to Synapse

To use Synapse, you'll need to
[register](https://www.synapse.org/#!RegisterAccount:0)
for an account. The Synapse website can authenticate using a Google account,
but you'll need to take the extra step of creating a Synapse password
to use the programmatic clients.

Once that's done, you'll be able to load the library and login:

```{r eval=F}
library(synapser)
synLogin('me@nowhere.com', 'secret')
```

You can also create a file `.synapseConfig` in your home directory containing login credentials:
```
[authentication]
username=me@nowhere.com
password=secret
```

after which you can log in without typing your credentials:
```{r collapse=TRUE}
library(synapser)
synLogin()
```

For more details see the native reference documentation:
```{r eval=F}
?synLogin
?synLogout
```

## Accessing Data

Synapse identifiers are used to refer to projects and data which are represented by 
*entity* objects. For example, the entity syn1899498 represents a tab-delimited file 
containing a 100 by 4 matrix. Getting the entity retrieves an object that holds metadata 
describing the matrix, and also downloads the file to a local cache:

```{r collapse=TRUE}
entity <- synGet('syn1899498')
```

View the entity's metadata in the R console:

```{r collapse=TRUE}
print(entity)
```

This is one simple way to read in a small matrix:

```{r collapse=TRUE}
read.table(entity$path)
```

View the entity in the browser:

```{r collapse=TRUE}
synOnweb('syn1899498')
```

For more details see the native reference documentation, e.g.:
```{r eval=F}
?synGet
?synOnweb
```


## Organizing data in a Project

You can create your own projects and upload your own data sets. Synapse stores
entities in a hierarchical or tree structure. Projects are at the top level and
must be uniquely named:

```{r collapse=TRUE}
# (We use a time stamp in this example just to help ensure uniqueness.)
projectName<-sprintf("My unique project created on %s", format(Sys.time(), "%a %b %d %H%M%OS4 %Y"))
project<-Project(projectName)
project<-synStore(project)
```

Creating a folder:
```{r collapse=TRUE}
dataFolder <- Folder('Data', parent=project)
dataFolder <- synStore(dataFolder)
```
Adding files to the project:
```{r collapse=TRUE}
filePath<- tempfile()
connection<-file(filePath)
writeChar("this is the content of the file", connection, eos=NULL)
close(connection)  
file <- File(path=filePath, parent=dataFolder)
file <- synStore(file)
```
