---
title: "Synapse R Client Overview"
author: "Bruce Hoff"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Synapse R Client Overview}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


## Overview

The `synapser` package provides an interface to
[Synapse](<http://www.synapse.org>), a collaborative
workspace for reproducible data intensive research projects,
providing support for:

* integrated presentation of data, code and text
* fine grained access control
* provenance_ tracking

The `synapser` package lets you communicate with the cloud-hosted
Synapse service to access data and create shared data analysis projects from
within Python scripts or at the interactive Python console. Other Synapse clients
exist for [Python](http://docs.synapse.org/python),
[Java](https://github.com/Sage-Bionetworks/Synapse-Repository-Services/tree/develop/client/synapseJavaClient>),
and [the web browser](https://www.synapse.org). 

If you're just getting started with Synapse,
have a look at the [Getting Started guides for Synapse](http://docs.synapse.org/articles/getting_started.html)
.

Good example projects are:

* [TCGA Pan-cancer (syn300013)](https://www.synapse.org/#!Synapse:syn300013)
* [Development of a Prognostic Model for Breast Cancer Survival in an Open Challenge Environment (syn1721874)](https://www.synapse.org/#!Synapse:syn1721874)
* [Demo projects (syn1899339)](https://www.synapse.org/#!Synapse:syn1899339)

## Installation

`synapser` is available as a ready-built package. It can
be installed or upgraded using the standard `install.packages()` command, adding the Sage Bionetworks repository to the repo list, e.g.:

```{r eval=F}
install.packages("synapser", repos=c("https://sage-bionetworks.github.io/drat", "https://cran.cnr.berkeley.edu/"))
```
Alternatively, configure your default repo's in your `~/.Rprofile` like so:
```{r eval=F}
options(repos=c("https://sage-bionetworks.github.io/drat", "https://cran.cnr.berkeley.edu/"))
```
after which you may run install.packages without specifying the repositories:
```{r eval=F}
install.packages("synapser")
```

## Connecting to Synapse

To use Synapse, you'll need to
[register](https://www.synapse.org/#!RegisterAccount:0)
for an account. The Synapse website can authenticate using a Google account,
but you'll need to take the extra step of creating a Synapse password
to use the programmatic clients.

Once that's done, you'll be able to load the library and login:

```{r eval=F}
library(synapser)
synLogin('me@nowhere.com', 'secret')
```

You can also create a file `.synapseConfig` in your home directory containing login credentials:
```
[authentication]
username=me@nowhere.com
password=secret
```

after which you can log in without typing your credentials:
```{r collapse=TRUE}
library(synapser)
synLogin()
```

For more details see the native reference documentation:
```{r eval=F}
?synLogin
?synLogout
```

## Accessing Data

Synapse identifiers are used to refer to projects and data which are represented by 
*entity* objects. For example, the entity syn1899498 represents a tab-delimited file 
containing a 100 by 4 matrix. Getting the entity retrieves an object that holds metadata 
describing the matrix, and also downloads the file to a local cache:

```{r collapse=TRUE}
entity <- synGet('syn1899498')
```

View the entity’s metadata in the R console:

```{r collapse=TRUE}
print(entity)
```

This is one simple way to read in a small matrix:

```{r collapse=TRUE}
read.table(entity$path)
```

View the entity in the browser:

```{r collapse=TRUE}
synOnweb('syn1899498')
```

For more details see the native reference documentation, e.g.:
```{r eval=F}
?synGet
?synOnweb
```


## Organizing data in a Project

You can create your own projects and upload your own data sets. Synapse stores
entities in a hierarchical or tree structure. Projects are at the top level and
must be uniquely named:

```{r collapse=TRUE}
# (We use a time stamp in this example just to help ensure uniqueness.)
projectName<-sprintf("My unique project created on %s", format(Sys.time(), "%a %b %d %H%M%OS4 %Y"))
project<-Project(projectName)
project<-synStore(project)
```

Creating a folder:
```{r collapse=TRUE}
dataFolder <- Folder('Data', parent=project)
dataFolder <- synStore(dataFolder)
```
Adding files to the project:
```{r collapse=TRUE}
filePath<- tempfile()
connection<-file(filePath)
writeChar("this is the content of the file", connection, eos=NULL)
close(connection)  
file <- File(path=filePath, parent=dataFolder)
file <- synStore(file)
```

You can print the properties of an entity (such as the file we just created):
```{r collapse=TRUE}
file$properties
```


TODO talk about the location of the downloaded file, how to change it, how the 
client keeps from downloading it twice.

Most other properties are immutable, but you *can* change an entity's name:
```{r collapse=TRUE}
file$properties$name<-"different name"
```

Update Synapse with the change:
```{r collapse=TRUE}
file<-synStore(file)
file$properties
```

Content can be deleted:
```{r collapse=TRUE}
synDelete(file)
```
Deletion of a project will also delete its contents, in this case the folder:
```{r collapse=TRUE}
folderId<-dataFolder$properties$id
synDelete(project)
tryCatch(
	synGet(folderId),
	error=function(e) {
		message(sprintf("Retrieving a deleted folder causes: %s", as.character(e)))
	},
	silent = TRUE
)
```


In addition to simple data storage, Synapse entities can be annotated with
key/value metadata, described in markdown documents (wikis), and linked
together in provenance graphs to create a reproducible record of a data
analysis pipeline.

For more details see the native reference documentation, e.g.:
```{r eval=F}
?Project
?Folder
?File
?Link
?synStore
```


## Annotating Synapse entities

```{r collapse=TRUE}
# (We use a time stamp just to help ensure uniqueness.)
projectName<-sprintf("My unique project created on %s", format(Sys.time(), "%a %b %d %H%M%OS4 %Y"))
project<-Project(projectName)
project<-synStore(project)
```

```{r collapse=TRUE}
synSetAnnotations(project, list(annotationName="annotationValue"))

project<-synGet(project$properties$id)

project$annotations

synGetAnnotations(project)
```

To add a new annotation to the ones currently defined for an entity:
TODO remove the 'eval=F' in the following block.  It causes the Windows build to fail.
```{r eval=F, collapse=TRUE}

annotations<-synGetAnnotations(project)

annotations[["numeric_annotation_name"]]<-42

annotations<-synSetAnnotations(project, annotations)

annotations
```

## Provenance

Synapse provides tools for tracking 'provenance', or the transformation of raw data
into processed results, by linking derived data objects to source data and the
code used to perform the transformation.


TODO example with Activity

```{r eval=F}
?Activity
```

## Tables
TODO

Tables can be built up by adding sets of rows that follow a user-defined schema
and queried using a SQL-like syntax.

```{r eval=F}
?Schema
?Column
?synGetColumns
?synGetTableColumns
```

## Wikis

TODO

Wiki pages can be attached to an Synapse entity (i.e. project, folder, file, etc).
Text and graphics can be composed in markdown and rendered in the web view of
the object.

```{r eval=F}
?synGetWiki
?Wiki
```

## Evaluations

An evaluation is a Synapse construct useful for building processing pipelines and
for scoring predictive modeling and data analysis challenges.

```{r eval=F}
?getEvaluation
?submit
?getSubmissions
?getSubmission
?getSubmissionStatus
```

## Sharing Access to Content

By default, data sets in Synapse are private to your user account, but they can
easily be shared with specific users, groups, or the public.

TODO

```{r eval=F}
?synGetPermissions
?synSetPermissions
```


```{r collapse=TRUE}
synDelete(project)
```

## File Views
TODO should this be in a separate vignette?


## Accessing the API directly

These methods enable access to the Synapse REST(ish) API taking care of details
like endpoints and authentication. See the
[REST API documentation](<http://docs.synapse.org/rest/).

```{r eval=F}
?synRestGET
?synRestPOST
?synRestPUT
?synRestDELETE
```


## Synapse Utilites
TODO

## More information

For more information see the
[Synapse User Guide](http://docs.synapse.org/articles/).

 
