---
title: "Tables"
author: "Bruce Hoff"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Tables}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
## Tables
Synapse Tables enable storage of tabular data in Synapse in a form that can be queried using a SQL-like query language.

A table has a Schema and holds a set of rows conforming to that schema.

A Schema defines a series of Column of the following types: STRING, DOUBLE, INTEGER, BOOLEAN, DATE, ENTITYID, FILEHANDLEID, LINK, LARGETEXT, USERID

Preliminaries:
```{r collapse=TRUE}
library(synapser)
synLogin()
# Create a new project
# use hex_digits to generate random string
hex_digits <- c(as.character(0:9), letters[1:6])
projectName <- sprintf("My unique project %s", paste0(sample(hex_digits, 32, replace = TRUE), collapse = ""))
project <- Project(projectName)
project <- synStore(project)
```

To create a Table, you first need to create a Table Schema. This defines the columns of the table:
```{r collapse=TRUE}
cols <- list(
    Column(name = "Name", columnType = "STRING", maximumSize = 20),
    Column(name = "Chromosome", columnType = "STRING", maximumSize = 20),
    Column(name = "Start", columnType = "INTEGER"),
    Column(name = "End", columnType = "INTEGER"),
    Column(name = "Strand", columnType = "STRING", enumValues = list("+", "-"), maximumSize = 1),
    Column(name = "TranscriptionFactor", columnType = "BOOLEAN"))

schema <- Schema(name = "My Favorite Genes", columns = cols, parent = project)
```

Next, let’s load some data. Let’s say we had a file, genes.csv:
```{r collapse=TRUE}
genes <- data.frame(
  Name = c("foo", "arg", "zap", "bah", "bnk", "xyz"),
  Chromosome = c(1, 2, 2, 1, 1, 1),
  Start = c(12345, 20001, 30033, 40444, 51234, 61234),
  End = c(126000, 20200, 30999, 41444, 54567, 68686),
  Strand = c("+", "+", "-", "-", "+", "+"),
  TranscriptionFactor = c(F, F, F, F, T, F))
```

Let’s store that in Synapse:
```{r collapse=TRUE}
table <- Table(schema, genes)
table <- synStore(table)
tableId <- table$schema$properties$id
```

The Table() function takes two arguments, a schema object and data in some form, which can be:

* a path to a CSV file
* a data frame
* a RowSet object
* a list of lists where each of the inner lists is a row

We now have a table populated with data. Let’s try to query:
```{r collapse=TRUE}
results <- synTableQuery(sprintf("select * from %s where Chromosome='1' and Start < 41000 and End > 20000", tableId))
```

`synTableQuery()` downloads the data and saves it to a csv file at location:
```{r collapse=TRUE, eval=F}
results$filepath
```

To load the data into an R data.frame:
```{r collapse=TRUE}
as.data.frame(results)
```

## Changing Data

Once the schema is settled, changes come in two flavors: appending new rows and updating existing ones.

Changing data in a table requires row IDs and version numbers for each row to be modified (called `ROW_ID` and `ROW_VERSION`). We get those by querying before updating. Minimizing change sets to contain only rows that actually change will make processing faster. Appending new rows can be accomplished by leaving values for `ROW_ID` and `ROW_VERSION` blank.

Appending new rows is fairly straightforward. To continue the previous example, we might add some new genes:
```{r collapse=TRUE}
moreGenes <- data.frame(
  Name = c("abc", "def"),
  Chromosome = c(2, 2),
  Start = c(12345, 20001),
  End = c(126000, 20200),
  Strand = c("+", "+"),
  TranscriptionFactor = c(F, F))
synStore(Table(tableId, moreGenes))
```

For example, let’s update the names of some of our favorite genes:
```{r collapse=TRUE}
results <- synTableQuery(sprintf("select * from %s where Chromosome='1'", tableId))
df <- as.data.frame(results)
df["Name"] <- c("rzing", "zing1", "zing2", "zing3")
```

Let’s save that:
```{r collapse=TRUE}
table <- Table(tableId, df)
table <- synStore(table)
```

Now, query the table again to see your changes:
```{r collapse=TRUE}
results <- synTableQuery(sprintf("select * from %s limit 10", tableId))
as.data.frame(results)
```

One other piece of information required for making changes to tables is the etag, which is used by the Synapse server to prevent concurrent users from making conflicting changes through a technique called optimistic concurrency. This comes as a result of running `synTableQuery`. In the example above, you could see the etag by running `results$etag` - but you should never have to use it directly. In case of a conflict, your update may be rejected. You then have to do another query and try your update again.

## Changing Table Structure

Adding columns can be done using the methods Schema$addColumn() or addColumns() on the Schema object:
```{r collapse=TRUE}
schema <- synGet(tableId)
newColumn <- synStore(Column(name = "Note", columnType = "STRING", maximumSize = 20))
schema$addColumn(newColumn)
schema <- synStore(schema)
```

Renaming or otherwise modifying a column involves removing the column and adding a new column:
```{r collapse=TRUE}
schema$removeColumn(newColumn)
notesColumn <- synStore(Column(name = "Notes", columnType = "STRING", maximumSize = 20))
schema$addColumn(notesColumn)
schema <- synStore(schema)
```

Now we can set the values for the new column:
```{r collapse=TRUE}
results <- synTableQuery(sprintf("SELECT * FROM %s", tableId))
data <- as.data.frame(results)
data["Notes"] <- c("check name", NA, NA, NA, "update test", NA, NA, NA)
synStore(Table(tableId, data))
```

## Updating Column Type

Column "Notes" has type STRING with "maximumSize" set to 20. We cannot add a new row with "Notes" as "a very looooooooong note" since it has more than 20 characters. Let"s change the ColumnType to "STRING" with "maximumSize" set to 100:
```{r collapse=TRUE}
# getting the existing table metadata and data
originalSchema <- synGet(tableId)
oldQueryResults <- synTableQuery(sprintf("SELECT * FROM %s", tableId))
oldData <- as.data.frame(oldQueryResults)

# remove the column
originalSchema$removeColumn(notesColumn)
newSchema <- synStore(originalSchema)

# create a new Column
newCol <- Column(name = "Notes", columnType = "STRING", maximumSize = 100)

# add the new column to the new table
newSchema$addColumn(newCol)
newSchema <- synStore(newSchema)

# copy the data over to the new column
newQueryResults <- synTableQuery(sprintf("SELECT * FROM %s", newSchema$properties$id))
newData <- as.data.frame(newQueryResults)
newData["Notes"] <- oldData["Notes"]

# save the change
synStore(Table(tableId, newData))

# add the new data
moreGenes <- data.frame(
    Name = c("not_sure"),
    Chromosome = c(2),
    Start = c(12345),
    End = c(126000),
    Strand = c("+"),
    TranscriptionFactor = c(F),
    Notes = c("a very looooooooong note"))
synStore(Table(tableId, moreGenes))
```

To access a column that you do not have a reference to, please see:
```{r collapse=TRUE, eval=F}
?synGetColumn
?synGetColumns
```

## Table Attached Files

Synapse tables support a special column type called ‘File’ which contain a file handle, an identifier of a file stored in Synapse. Here’s an example of how to upload files into Synapse, associate them with a table and read them back later:

```{r collapse=TRUE, eval=F}
newCols <- list(
    Column(name = "artist", columnType = "STRING", maximumSize = 50),
    Column(name = "album", columnType = "STRING", maximumSize = 50),
    Column(name = "year", columnType = "INTEGER"),
    Column(name = "catalog", columnType = "STRING", maximumSize = 50),
    Column(name = "cover", columnType = "FILEHANDLEID"))
newSchema <- synStore(Schema(name = "Jazz Albums", columns = newCols, parent = project))

newData <- data.frame(
  artist = c("John Coltrane", "Sonny Rollins", "Sonny Rollins", "Kenny Burrel"),
  album = c("Blue Train", "Vol. 2", "Newk's Time", "Kenny Burrel"),
  year = c(1957, 1957, 1958, 1956),
  catalog = c("BLP 1577", "BLP 1558", "BLP 4001", "BLP 1543")
)

# writing some temp files to upload or pointing to existing files in your system

files <- c("coltraneBlueTrain.jpg", "rollinsBN1558.jpg", "rollinsBN4001.jpg", "burrellWarholBN1543.jpg")

# upload to filehandle service
files <- lapply(files, function (f) {
  temp_file <- tempfile(f)
  write(f, file = temp_file)
  synUploadSynapseManagedFileHandle(f)
  })

# get the filehandle ids
fileHandleIds <- sapply(files, function(f) f[1]$id)
newData["covers"] <- fileHandleIds

newTable <- synStore(Table(newSchema$properties$id, newData))
```
To download attached files in a table:
```{r collapse=TRUE, eval=F}
result <- synTableQuery(sprintf("select * from %s", newTable$schema$properties$id))
data <- synDownloadTableColumns(result, columns = c("covers"))
```

## Set Annotations
A table schema is a Synapse entity. Annotations on table works the same way as annotations on any other entity types.

To set annotation on table, use synSetAnnotations() on the schema:
```{r collapse=TRUE}
schema <- synGet(tableId)
synSetAnnotations(schema, list(category = "test"))
```
To view annotations on table, retrieve the schema:
```{r collapse=TRUE}
schema <- synGet(tableId)
schema$annotations
```

Please visit [synapser vignettes](synapser.html#annotating-synapse-entities) to read more about annotations.

## Deleting Rows
Query for the rows you want to delete and call synDelete on the results:
```{r collapse=TRUE}
results <- synTableQuery(sprintf("select * from %s where Chromosome='2'", tableId))
deleted <- synDelete(results$asRowSet())
```

## Deleting Table
Deleting the schema deletes the whole table and all rows:
```{r collapse=TRUE}
synDelete(schema)
```

## Queries
The query language is quite similar to SQL select statements, except that joins are not supported. The documentation for the Synapse API has lots of [query examples](http://docs.synapse.org/rest/org/sagebionetworks/repo/web/controller/TableExamples.html).

For more details see the native reference documentation, e.g.:

```{r eval=F}
?Schema
?Column
?Row
?Table
```

```{r collapse=TRUE}
synDelete(project)
```
