---
title: "Tables"
author: "Bruce Hoff"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Tables}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
## Tables
Synapse Tables enable storage of tabular data in Synapse in a form that can be queried using a SQL-like query language.

A table has a Schema and holds a set of rows conforming to that schema.

A Schema defines a series of Column of the following types: STRING, DOUBLE, INTEGER, BOOLEAN, DATE, ENTITYID, FILEHANDLEID, LINK, LARGETEXT, USERID

Preliminaries:
```{r collapse=TRUE}
library(synapser)
synLogin()
# Create a new project
projectName<-sprintf("My unique project created on %s", format(Sys.time(), "%a %b %d %H%M%OS4 %Y"))
project<-Project(projectName)
project<-synStore(project)

```
To create a Table, you first need to create a Table Schema. This defines the columns of the table:
```{r collapse=TRUE}
cols <- list(
    Column(name='Name', columnType='STRING', maximumSize=20),
    Column(name='Chromosome', columnType='STRING', maximumSize=20),
    Column(name='Start', columnType='INTEGER'),
    Column(name='End', columnType='INTEGER'),
    Column(name='Strand', columnType='STRING', enumValues=list('+', '-'), maximumSize=1),
    Column(name='TranscriptionFactor', columnType='BOOLEAN'))

schema <- Schema(name='My Favorite Genes', columns=cols, parent=project)
```
Next, let’s load some data. Let’s say we had a file, genes.csv:

```{r collapse=TRUE}
genes<-data.frame(
	Name=c("foo", "arg", "zap", "bah", "bnk", "xyz"), 
	Chromosome=c(1,2,2,1,1,1), 
	Start=c(12345,20001,30033,40444,51234,61234),
	End=c(126000,20200,30999,41444,54567,68686),
	Strand=c('+', '+', '-', '-', '+', '+'),
	TranscriptionFactor=c(F,F,F,F,T,F))
genesFile<-tempfile()
write.csv(genes, file=genesFile, row.names=FALSE)
```

Let’s store that in Synapse:
```{r collapse=TRUE}
table<-Table(schema, genesFile)
table<-synStore(table)
```

The Table() function takes two arguments, a schema object and data in some form, which can be:
* a path to a CSV file
* a data frame
* a RowSet object
* a list of lists where each of the inner lists is a row

We now have a table populated with data. Let’s try to query:

```{r collapse=TRUE, eval=F}
results <- synTableQuery(sprintf("select * from %s where Chromosome='1' and Start < 41000 and End > 20000", table$schema$properties$id))
for (row in results$asDataFrame()) {
	print(row)
}
```

# continue at 'Pandas' section here: http://docs.synapse.org/python/Table.html#module-synapseclient.table

TODO make sure this is at the bottom of the vignette
```{r collapse=TRUE}
synDelete(project)
```
