---
title: "Tables"
author: "Bruce Hoff"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Tables}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
## Tables
Synapse Tables enable storage of tabular data in Synapse in a form that can be queried using a SQL-like query language.

A table has a Schema and holds a set of rows conforming to that schema.

A Schema defines a series of Column of the following types: STRING, DOUBLE, INTEGER, BOOLEAN, DATE, ENTITYID, FILEHANDLEID, LINK, LARGETEXT, USERID

Preliminaries:
```{r collapse=TRUE}
library(synapser)
synLogin()
# Create a new project
projectName<-sprintf("My unique project created on %s", format(Sys.time(), "%a %b %d %H%M%OS4 %Y"))
project<-Project(projectName)
project<-synStore(project)

```
To create a Table, you first need to create a Table Schema. This defines the columns of the table:
```{r collapse=TRUE}
cols <- list(
    Column(name='Name', columnType='STRING', maximumSize=20),
    Column(name='Chromosome', columnType='STRING', maximumSize=20),
    Column(name='Start', columnType='INTEGER'),
    Column(name='End', columnType='INTEGER'),
    Column(name='Strand', columnType='STRING', enumValues=list('+', '-'), maximumSize=1),
    Column(name='TranscriptionFactor', columnType='BOOLEAN'))

schema <- Schema(name='My Favorite Genes', columns=cols, parent=project)
```
Next, let’s load some data. Let’s say we had a file, genes.csv:

```{r collapse=TRUE}
genes<-data.frame(
	Name=c("foo", "arg", "zap", "bah", "bnk", "xyz"), 
	Chromosome=c(1,2,2,1,1,1), 
	Start=c(12345,20001,30033,40444,51234,61234),
	End=c(126000,20200,30999,41444,54567,68686),
	Strand=c('+', '+', '-', '-', '+', '+'),
	TranscriptionFactor=c(F,F,F,F,T,F))
genesFile<-tempfile()
write.csv(genes, file=genesFile, row.names=FALSE)
```

Let’s store that in Synapse:
```{r collapse=TRUE}
table<-Table(schema, genesFile)
table<-synStore(table)
tableId<-table$schema$properties$id
```

The Table() function takes two arguments, a schema object and data in some form, which can be:

* a path to a CSV file
* a data frame
* a RowSet object
* a list of lists where each of the inner lists is a row

We now have a table populated with data. Let’s try to query:

```{r collapse=TRUE}
results <- synTableQuery(sprintf("select * from %s where Chromosome='1' and Start < 41000 and End > 20000", tableId))
results$asDataFrame()
```

## Changing Data

Once the schema is settled, changes come in two flavors: appending new rows and updating existing ones.

Appending new rows is fairly straightforward. To continue the previous example, we might add some new genes:
```{r collapse=TRUE}
more_genes<-data.frame(
	Name=c("abc", "def"), 
	Chromosome=c(2,2), 
	Start=c(12345,20001),
	End=c(126000,20200),
	Strand=c('+', '+'),
	TranscriptionFactor=c(F,F))
more_genes_file<-tempfile()
write.csv(more_genes, file=more_genes_file, row.names=FALSE)
synStore(Table(tableId, more_genes_file))
```

Updating rows requires an etag, which identifies the most recent change set plus row IDs and version numbers for each row to be modified. We get those by querying before updating. Minimizing changesets to contain only rows that actually change will make processing faster.

For example, let’s update the names of some of our favorite genes:
```{r collapse=TRUE}
results <- synTableQuery(sprintf("select * from %s where Chromosome='1'", tableId))
df <-results$asDataFrame(rowIdAndVersionInIndex=FALSE)
df['Name'] = c('rzing', 'zing1', 'zing2', 'zing3')
write.csv(df, results$filepath, row.names=FALSE)
```

Let’s save that:
```{r collapse=TRUE}
table<-Table(tableId, results$filepath)
table<-synStore(table)
```

Now, query the table again to see your changes:
```{r collapse=TRUE}
results <- synTableQuery(sprintf("select * from %s limit 10", tableId))
results$asDataFrame()
```
The etag is used by the server to prevent concurrent users from making conflicting changes, a technique called optimistic concurrency. In case of a conflict, your update may be rejected. You then have to do another query an try your update again.

http://docs.synapse.org/python/Table.html#module-synapseclient.table

TODO make sure this is at the bottom of the vignette
```{r collapse=TRUE}
synDelete(project)
```
