---
title: "Troubleshooting"
author: "KimYen Ladia"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Troubleshooting}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
## Troubleshooting

For users who are using a Windows machine and a network mapped drive, where the folder is mapped from a drive, e.g., `H:\\Documents\\R\\win-library\\3.4` to a network location, e.g., `\\\\samba-xxx/home/<your_home_dir>/Documents/R/win-library/3.4`, you may experience the following error:
```{r eval = FALSE}
> library(synapser)

Loading required package: PythonEmbedInR

Error: package or namespace load failed for ‘PythonEmbedInR’:

.onLoad failed in loadNamespace() for 'PythonEmbedInR', details:

  call: library.dynam.unload("PythonEmbedInR", packageRootDir)

  error: DLL ‘PythonEmbedInR.dll’ was not loaded

Error: package ‘PythonEmbedInR’ could not be loaded
```

First, verify that is R, `.libPaths()` uses the non-Windows reference to the network file share, for example:
```{r eval = FALSE}
> .libPaths()
[1] "\\\\samba-xxx/home/<your_home_dir>/Documents/R/win-library/3.4" "\\\\samba-xxx/home/<your_home_dir>/Documents/R/R-3.4.4/library"
```

Then update `.libPaths` to have the Windows-style reference, for example:
```{r eval = FALSE}
> .libPaths("H:\\Documents\\R\\win-library\\3.4")
> .libPaths()
[1] "H:/Documents/R/win-library/3.4" "\\\\samba-xxxo/home/<your_home_dir>/Documents/R/R-3.4.4/library"

```

Now, you should be able to load `synapser` without errors.
