---
title: "Troubleshooting"
author: "Tom Yu"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Troubleshooting}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
## Troubleshooting

### Build Synapser
When You install Synapser, You may encounter a similar error:

```{console}
$ R CMD BUILD .

Execution halted
ERROR: configuration failed for package ‘synapser’
* removing ‘/private/var/folders/2n/xwh21m9s3sq_mhnkbc6n6z_c0000gn/T/RtmpZ47UPP/Rinstf55478c3f09/synapser’
```

For this error, You can try first remove $HOME\.virtualenvs then run the build command again.

### Build Synapser when creating vignettes
When you build synapser, you may encounter the following error:

```{console}
$ R CMD BUILD .

Error: processing vignette 'upload.Rmd' failed with diagnostics:
No credentials provided.
--- failed re-building ‘upload.Rmd’
```

First, create a file named `.synapseConfig` in your home directory. Then add the following lines to the file:

```{console}
[authentication]
username = your_username
password = your_password
```

Now, you should be able to build Synapser.

### Using Synapser with reticulate

Synapser is compatible with [reticulate](https://rstudio.github.io/reticulate/) and the two packages can be used together in the same R session.

As a workaround, if you wish to communicate with Synapse in an R session that also uses reticulate, you can use the [Synapse Python client](https://python-docs.synapse.org/build/html/index.html) through reticulate.


### Type Conversions

The core of the Synapser library is the [python client](https://python-docs.synapse.org/). [Reticulate](https://rstudio.github.io/reticulate/) is used to translate R to Python and back.

These are the documented conversions from R to  Python: [Reticulate Type Conversions](https://rstudio.github.io/reticulate/articles/calling_python.html#type-conversions)

`NA` values in R do not have a direct equivalent in Python. This has been a topic of great discussion. Synapser is not trying to solve this. `NA` values in R are handled by Reticulate per the implementation defined in Reticulate.

When working with R and Reticulate, NA values in types tables are converted by Reticulate depending on the data type.

For numeric data types, NA values are converted to the corresponding R value `NA_real_`.
For integer data types, NA values are converted to `NA_integer_`.
For logical data types, NA values are converted to `NA_logical_`.
For character data types, NA values are converted to `NA_character_`.

In addition, Reticulate also converts any values that cannot be coerced to the specified data type to `NA_character_`.

Reticulate performs these conversions automatically.

When Reticulate converts R objects to Python objects, the NA values in R types tables are converted to Python's `None`.

R's `NA` is a logical constant of length 1 which is used to represent the absence of a value. In Python, the equivalent of `NA` is `None`, which is a special constant used to indicate the absence of a value.

Reticulate automatically converts `NA` values in R types tables to `None` when converting R objects to Python. This ensures that the data types and values remain consistent between the two languages.

Examples:

* Reticulate converts logical `NA`s to Python `True`, which convert back to `TRUE` in R
```{r collapse = TRUE}
na_logical = NA
class(na_logical)
na_py <- reticulate::r_to_py(na_logical)
na_py
na_r <- reticulate::py_to_r(na_py)
na_r
```

```{r collapse = TRUE}
na_logical = c(T, F, NA)
class(na_logical)
na_py <- reticulate::r_to_py(na_logical)
na_py
na_r <- reticulate::py_to_r(na_py)
na_r
```

* Reticulate converts character `NA`s to Python strings, which convert back to characters in R
```{r collapse = TRUE}
na_char = c("T", "F", NA)
class(na_char)
na_py <- reticulate::r_to_py(na_char)
na_py
na_r <- reticulate::py_to_r(na_py)
na_r
```

* Reticulate converts numeric `NA`s to Python `np.nan`, which convert back to `NA` in R
```{r collapse = TRUE}
na_num = c(1538006762583, 1538006762584, NA)
class(na_num)
na_py <- reticulate::r_to_py(na_num)
na_py
na_r <- reticulate::py_to_r(na_py)
na_r
```
