---
title: "File Upload"
author: "KimYen Ladia"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{File Upload}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
## File Upload

Files in Synapse are versionable. Please see [Files and Versioning](https://docs.synapse.org/articles/files_and_versioning.html) for more information about how versions in Files works.

Preliminaries:
```{r collapse=TRUE}
library(synapser)
synLogin()
# Create a new project
# use hex_digits to generate random string
hex_digits <- c(as.character(0:9), letters[1:6])
projectName <- sprintf("My unique project %s", paste0(sample(hex_digits, 32, replace = TRUE), collapse = ""))
project <- Project(projectName)
project <- synStore(project)

# Create a file
file_path <- tempfile("testUpload.txt")
connection <- file(file_path)
writeChar("this is the content of the file", connection, eos = NULL)
close(connection)
file <- File(path = file_path, parent = project)
file <- synStore(file)
file_id <- file$properties$id
```

### Uploading a New Version

Uploading a new version follows the same steps as uploading a file for the first time - use the same file name and store it in the same location (e.g., the same parentId). It is recommended to add a comment to the new version in order to easily track differences at a glance. The example file raw_data.txt will now have a version of 2 and a comment describing the change.

*Explicit example:*
```{r collapse=TRUE}
# fetch the file in Synapse
file_to_update <- synGet(file_id, downloadFile = FALSE)

# create the second version as a separated file
file_path <- tempfile("testUpload.txt")
connection <- file(file_path)
writeChar("this is version 2", connection, eos = NULL)
close(connection)

# save the local path to the new version of the file
file_to_update$path <- file_path

# add a version comment
file_to_update$versionComment <- 'change the file content'

# store the new file
updated_file <- synStore(file_to_update)
```

*Implicit example:*
```{r collapse=TRUE}
# To create a new version of that file, make sure you store it with the exact same name
new_file <- synStore(File(file_path,  parentId = project$properties$id))
```

### Updating Annotations or Provenance without Changing Versions

Any change to a File will automatically update its version. If this isn’t the desired behavior, such as minor changes to the metadata, you can set forceVersion=FALSE with the Python client. For command line, the commands set-annotations and set-provenance will update the metadata without creating a new version. Adding/updating annotations and provenance in the web client will also not cause a version change.

*Important: Because Provenance is tracked by version, set `forceVersion=FALSE` for minor changes to avoid breaking Provenance.*

### Setting Annotations without Changing Version

```{r collapse=TRUE}
# Get file from Synapse, set downloadFile = FALSE since we are only updating annotations
file <- synGet(file_id, downloadFile = FALSE)

# Add annotations
file$annotations = c("fileType" = "bam", "assay" = "RNA-seq")

# Store the file without creating a new version
file <- synStore(file, forceVersion = FALSE)
```

### Setting Provenance without Changing Version

To set Provenance without changing the file version:

```{r collapse=TRUE}
# Get file from Synapse, set downloadFile = FALSE since we are only updating provenance
file <- synGet(file_id, downloadFile = FALSE)

# Store the file without creating a new version
file <- synStore(file,  activity = Activity(used = project$properties$id), forceVersion = FALSE)
```

### Downloading a Specific Version

By default, the File downloaded will always be the most recent version. However, a specific version can be downloaded by passing the version parameter:

```{r collapse=TRUE}
entity <- synGet(file_id, version = 1)
```

```{r collapse=TRUE}
synDelete(project)
```
